opponent modelling

### 2019-11-01 comp_model_1019.Rmd

```{r}
dat <- read.csv("data20180719.csv")
#dat <- read.csv("test_data.csv")
```

@@

```{r}
# using some functions from the 'tidyverse' (also for me to get use to them ;-)
library(tidyr)
library(dplyr)
# transform 'winner' variable in numeric score
dat$score <- recode(dat$winner, human = 1, tie = 0, ai = -1)
# create a new variable 'block' with round 1...25 = block 1 and round 26...50 as block 2
dat$block <- as.numeric(cut(dat$round,2))
# create a new data.frame in a 'wide format'
widedata <- dat %>% 
  unite(game_block,game,block) %>% # this creates a new variable which combines game and block
    group_by(human_id,condition,game_block) %>% # let the functions know you want to separate things by ID, condition, and game_block
      summarize(mean_score = mean(score)) %>% # compute average score (i.e wins - losses)
        spread(game_block,mean_score) # reformat in the 'wide' format
# save the data file as a .csv to use in e.g. SPSS
write.csv(widedata,row.names=FALSE,file="scores_wide.csv")

@@

naive_bayes_modelling <- list()
for(id in unique(dat$human_id)) {
  naive_bayes_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  naive_bayes_modelling[[id]] <- optim(c(0.1,0.1),fn=naive_bayes,gr = NULL, data=tdat,opp_strategy_vec = c("level0","level1","level2"),"-2loglik", lower = c(0.01,0.01), upper = c(0.99,0.99), method="L-BFGS-B")
}

save(naive_bayes_modelling,file="naive_bayes_modelling.RData")

```
```{r}
Nbayes_par_results <- data.frame()
for(id in unique(dat$human_id)) {
Nbayes_par_results <- rbind(Nbayes_par_results,
                       data.frame(
                         "id" = id,
                         "condition" = dat[dat$human_id==id,"condition"][1],
                         # Theta is the parameter governing AI stochasticity. Truth is 0.9
                         "theta" = naive_bayes_modelling[[id]]$par[1],
                         # Epsilone is parameter showing cases where human deviates from predictions about opponent (lower better)
                         "eps" = naive_bayes_modelling[[id]]$par[2]))
}

# Plotting the theta/eps scatter for all participants...
plot(Nbayes_par_results$theta,Nbayes_par_results$eps,xlab="Theta", ylab= "Epsilone",col = ifelse(as.character(Nbayes_par_results$condition) == "Level2",'red','green'))

write.csv(Nbayes_par_results,file="Nbayes_par_results.csv",row.names = FALSE)
#kable(Nbayes_par_results)
```

## Q-learning
```{r}
# 
# Q_learn <- function(beta,lambda,data){
#   # actions 
#   # beta= inverse temperature parameter in softmax choicd=e function
#   # lambda = learning rate (one per game?)
#   # data : participant and ai choice data. 
#   # Returns Q-values per trial and predicts choice using softmax
#   Q_vals_RPS = rep(0.0,3)
#   names(Q_vals_RPS) <- c("R","P","S")
#   Q_vals_FWG = rep(0.0,3)
#   names(Q_vals_FWG) <- c("F","W","G")
#   Q_vals_NUM = rep(0.0,5)
#   names(Q_vals_NUM) <- c("1","2","3","4","5")
#   
#   for(t in 1:nrow(data)) {
#     t_game <- data[t,"game"]
#     Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, numbers = Q_vals_NUM)
#     if(t_game == "numbers") nopts <- 5 else nopts <- 3
#     if(data[t,"round"] == 1) {
#       # first round is uniform prediction
#       Q_vals <- rep(0.0, nopts)
#       #lik[,t] <- prior_vec
#     } else {
#       
#     }
#     
#     
#   }
#   
# }

@@

>>>>

```{r}
dat <- read.csv("data20180719.csv")
# dat <- read.csv("test_data.csv")
```

@@

```{r}
# using some functions from the 'tidyverse' (also for me to get use to them ;-)
library(tidyr)
library(dplyr)
# transform 'winner' variable in numeric score
dat$score <- recode(dat$winner, human = 1, tie = 0, ai = -1)
# create a new variable 'block' with round 1...25 = block 1 and round 26...50 as block 2
dat$block <- as.numeric(cut(dat$round,2))
# create a new data.frame in a 'wide format'
# widedata <- dat %>%
#   unite(game_block,game,block) %>% # this creates a new variable which combines game and block
#     group_by(human_id,condition,game_block) %>% # let the functions know you want to separate things by ID, condition, and game_block
#       summarize(mean_score = mean(score)) %>% # compute average score (i.e wins - losses)
#         spread(game_block,mean_score) # reformat in the 'wide' format
# # save the data file as a .csv to use in e.g. SPSS
# write.csv(widedata,row.names=FALSE,file="scores_wide.csv")

@@

```{r}
# 
# naive_bayes_modelling <- list()
# for(id in unique(dat$human_id)) {
#   naive_bayes_modelling[[id]] <- list()
#   tdat <- subset(dat,human_id == id)
#   naive_bayes_modelling[[id]] <- optim(c(0.1,0.1),fn=naive_bayes,gr = NULL, data=tdat,opp_strategy_vec = c("level0","level1","level2"),"-2loglik", lower = c(0.01,0.01), upper = c(0.99,0.99), method="L-BFGS-B")
# }
# 
# save(naive_bayes_modelling,file="naive_bayes_modelling.RData")
# 
# ```
# ```{r}
# Nbayes_par_results <- data.frame()
# for(id in unique(dat$human_id)) {
# Nbayes_par_results <- rbind(Nbayes_par_results,
#                        data.frame(
#                          "id" = id,
#                          "condition" = dat[dat$human_id==id,"condition"][1],
#                          # Theta is the parameter governing AI stochasticity. Truth is 0.9
#                          "theta" = naive_bayes_modelling[[id]]$par[1],
#                          # Epsilone is parameter showing cases where human deviates from predictions about opponent (lower better)
#                          "eps" = naive_bayes_modelling[[id]]$par[2]))
# }
# 
# # Plotting the theta/eps scatter for all participants...
# plot(Nbayes_par_results$theta,Nbayes_par_results$eps,xlab="Theta", ylab= "Epsilone",col = ifelse(as.character(Nbayes_par_results$condition) == "Level2",'red','green'))
# 
# write.csv(Nbayes_par_results,file="Nbayes_par_results.csv",row.names = FALSE)
#kable(Nbayes_par_results)
```

## Q-learning
```{r}

Q_learn <- function(par,data,return_value=c("-2loglik","likelihood_by_trial")){
  # actions
  # beta= inverse temperature parameter in softmax choice function
  # lambda = learning rate (one per game?)
  # data : participant and ai choice data.
  # Returns Q-values per trial and predicts choice using softmax
  beta <- par[1]
  lambda <- par[2]
  Q_vals_RPS = matrix(0.0,3)
  names(Q_vals_RPS) <- c("R","P","S")
  Q_vals_FWG = matrix(0.0,3)
  names(Q_vals_FWG) <- c("F","W","G")
  Q_vals_NUM = matrix(0.0,5)
  names(Q_vals_NUM) <- c("1","2","3","4","5")

  lik_hum <- matrix(0.0,nrow(data))

  for(t in 1:nrow(data)) {
    t_game <- data[t,"game"]
    if(t_game == "numbers") nopts <- 5 else nopts <- 3
    if(data[t,"round"] == 1) {
      # first round is uniform prediction
      Q_vals <- switch(as.character(t_game),rps=Q_vals_RPS,fwg = Q_vals_FWG, numbers = Q_vals_NUM)
      lik_hum[t] <- 1/nopts
    } else {
      # Get reward and past human action
      reward <- as.numeric(data[t-1,"score"])
      h_act_prev <- as.character(data[t-1,"h_action"])
      Q_vals[h_act_prev] <- Q_vals[[h_act_prev]] + as.numeric(lambda)*( reward - Q_vals[[h_act_prev]])

      # cat(Q_vals,"\n")

      # Assume human chooses action probabilistically using softmax on Q values
      probs <- exp(Q_vals/beta)/sum(exp(Q_vals/beta))

      # Get actual human action and compute likelihood
      h_act <- as.character(data[t,"h_action"])
      act_index <- match(h_act, names(Q_vals))
      lik_hum[t] <- probs[[act_index]]
    }
  }
  if(return_value == "-2loglik") {
    ret <- -2*sum(log(lik_hum))
    if(is.infinite(ret)) {
      return(1e+300)
    } else {
      return(ret)
    }
  }
  if(return_value == "likelihood_by_trial") return(lik_hum)

}

@@

```{r}
QL_modelling <- list()
for(id in unique(dat$human_id)) {
  QL_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  QL_modelling[[id]] <- optim(c(0.1,0.1),fn=Q_learn,gr = NULL, data=tdat,"-2loglik", lower = c(0.001,0.01), upper = c(0.99,10), method="L-BFGS-B")
}

save(QL_modelling,file="QL_modelling.RData")

## Q-learning with state space including last actions
